SAT Solving for Vesicle Traffic Systems in Cells

### 2017-06-22 paper/encoding.tex
more formulas

\begin{df}
  A VTS $G$ is tuple $(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$, where
  \begin{itemize}
  \item $\nodes$ is a set of nodes representing compartments in the VTS,
  \item $\mols$ is the set of molecules flowing in the network, 
  \item $\nlabel : \nodes \maps \powerset{\mols}$ defines the molecules present in the nodes  
  \item $ \nodes \times (\powerset{\mols}-\emptyset) \times \nodes \subseteq \edges$ is the set of edges
    with molecule sets as labels
  \item $\pairs : \mols \times \mols$ is pairing relating
  \item $\nodef : \mols \maps \powerset{\mols} \maps \booleans$ is activity maps for nodes
  \item $\edgef : \mols \maps \powerset{\mols} \maps \booleans $ is activity maps for edges
  \end{itemize}
\end{df}
$\pairs$ defines which molecules can fuse with which molecules.
%
Let $\pairs(M')$ denote $\{m|(m,m') \in P \text{ and } m' \in M'\}$.
%
$\nodef$ and $\edgef$ are used to define activity of molecules on
nodes and edges.
%
A molecule $k$ is {\em active} at node $n$ if $k \in \nlabel(n)$ and
$\nodef(k,\nlabel(n))$ is true.
%
A molecule $k$ is {\em active} at edges $(n,M',n')$ if $k \in M'$ and
$\edgef(k,M')$ is true.

@@

\item $n_{i,m}$ indicates if $m \in \nlabel(i)$
\item $e_{i,j,q}$ indicates if $q$th edge exists between $i$ and $j$.
\item $e_{i,j,q,m}$ indicates if $q$th edge between $i$ and $j$ contains molecule $m$.

\item $p_{k,k'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}

We also have auxiliary variables that will help us encode the properties 
\begin{enumerate}
\item dump = d{i,j,q}
      to represent drop of edge{i,j,q} from the graph we will use d{i,j,q}.

\item active\_node = a{i,k} 
      kth molecule is active on the node i.

\item presence\_edge = e{i,j,q,k}
      There is an qth edge between node i and j with kth molecule present.

\item active\_edge = b{i,j,q,k}
      There is an qth edge between node i and j with kth molecule active.

\item reachability = r{i,j,k,z}
      Node i and j are reachable with kth molecule present in z steps.

\end{enumerate}

Every variable is a Boolean bit, for example, n(0,1) represents node 0
with type 1 molecule present. Assignment to the Boolean variable
n(0,1) to True will mean that at node 0 type 1 molecule is present.

Basic notation for the formula we are trying to generate.
\begin{enumerate}

\item edge = e(i,j,q)      
      we will assign e(i,j,q) for a qth edge between node i and j.

\item dump = d(i,j,q)
      to represent drop of edge(i,j,q) from the graph we will use d(i,j,q).

\item presence node = n(i,k) 
	  kth molecule is present on the node i.

\item active\_node = a(i,k) 
      kth molecule is active on the node i.

\item presence\_edge = e(i,j,q,k)
      There is an qth edge between node i and j with kth molecule present.

\item active\_edge = b(i,j,q,k)
      There is an qth edge between node i and j with kth molecule active.

\item reachability = r(i,j,k,z)
      Node i and j are reachable with kth molecule present in z steps.

\item pairing matrix = p(k,k')
      Molecule k and k' are pairing molecules in the matrix.

\item Type of the Boolean function = sorts (takes k-1 arguments Bool and return Bool) 
\item Boolean function on nodes Function (fn\_{m},*sorts)

\item Boolean function on edge Function(fe\_{m},*sorts)

\end{enumerate}

\subsubsection{Regulation on nodes and edges}
Regulation of the molecules is based on the kind of the model under consideration. For example, model 1 (-V 1) has two different arbitrary Boolean functions dictating the activity of the molecules on edge and nodes. For variation 3 (-V 3) activity on edge is driven by pairing inhibition.

@@

A\_nn: No regulation on nodes i.e all the present molecules on the nodes are active.
\[ \bigwedge\limits_{i,k} n_{i,k} == a_{i,k} \, \]

A\_nb: Boolean regulation on nodes. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,k} n_{i,k} \supset a_{i,k} ==   f_n[k] (\bigvee_{ k' != k} n_{i,k'}) \, \]

A\_en: No regulation on edges i.e all the present molecules on the edges are active.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} == b_{i,j,q,k} \, \]

A\_eb: Boolean regulation on edges. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \supset b_{i,j,q,k} == f_e[k](\bigvee_{k' != k} e_{i,j,q,k'}) \, \]

A\_ep: Pairing inhibition. The activity of the molecules is defined by inhibition by other molecules based on the pairing matrix. 
\[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \supset e_{i,j,q,k'})] ) \supset \neg b_{i,j,q,k}) \, \]

\[ \bigwedge\limits_{i,j, q, k}  (e_{i,j,q,k} \land \neg [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \supset e_{i,j,q,k'})] ) \supset b_{i,j,q,k}) \, \]

\subsubsection{Constraints on the structure of the edges, nodes and pairing matrix}
A few additional constraints are required on the nodes, edges and pairing matrix to set up the essential conditions. We use Boolean variables C0-C5 to build all these constraints.

C0: The edge labels are subset of the node label of source compartment.

\[ \bigwedge\limits_{i,j,k} e_{i,j,k} \supset n_{i,k} \, \]

C1: The edge labels are subset of the node label of target compartment.

\[ \bigwedge\limits_{i,q}  e_{i,j,k} \supset n_{j,k} \, \]  
%C1 =  e_ijk -> aik and e_ijk -> ajk

C2: Self edges are not allowed.

\[  \bigwedge\limits_{i,q} \neg e_{i,i,q} \, \]

C4: Condition on p\_kk'. Diagonal blocks should be all 0's.

\[ \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y) \, \]

C5: Activity on the node. A molecule should be present to be active on a node.  
\[ \bigwedge\limits_{i,k} a_{i,k} \supset n_{j,k} \, \]

% - F0: b(i,j,k) $\supset$ e(i,j,k)

\subsubsection{Constraints on presence and activity of the molecule}
Few additional constraints that are required for the molecules presence and activity is build using F0 and F1. \newline

F0: An edge has to have one present molecule. F0 looks redundant.

\[ \bigwedge\limits_{i,j,q} \bigvee_k e_{i,j,q,k} \supset e_{i,j,q} \, \]

F1: If molecule is active on an edge then it should be present on the edge.

\[ \bigwedge\limits_{i,j,q,k} b_{i,j,q,k} \supset e_{i,j,q,k}\, \]

@@

% I have implemented: Changed the implementation in code too.
%\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \supset (r_{j,i,k,0}  \lor r_{j,i,k,1} \lor ... \lor r_{j,i,k,p}) \, \]

% Update: Better 
\[ \bigwedge\limits_{i,j,k} (\bigvee_{q} e_{i,j,q,k}) \supset (r_{j,i,k,0}  \lor r_{j,i,k,1} \lor ... \lor r_{j,i,k,p}) \, \]

@@

\[ \bigwedge\limits_{i,j,k,p} r_{i,j,k,p} \supset (\bigvee_{q} \, e_{i,j,q,k} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,k}) \land r_{i^{\prime},j,k,p-1} ) \, \]

\subsubsection{Fusion rule constraints}

Fusion rules consist of two different mechanisms.

\begin{enumerate}
\item  A \textbf{pairing mechanism} which determines compatible Q-R pairs on vesicles and compartments that can cause fusion.
\item \textbf{regulatory mechanisms} on the edges and nodes (possibly distinct) which regulates molecules activity based on the presence/absence of other molecules on the corresponding node or edge.
\end{enumerate}

Boolean variable F4 and F5 are used to model these constraints.

F4: For an edge to be valid, at least one SNARE pair on the vesicle and target compartment must be active and have a non-zero entry in the pairing matrix.

\[ \bigwedge\limits_{i,j,q} e_{i,j,q} \supset \bigvee_{k,k^{\prime}} (b_{i,j,q,k} \land a_{j,k^{\prime}} \land p_{k,k^{\prime}}) \, \]

F5: To ensure that fusion respects the graph structure by the edge under consideration, it should not be possible to fuse with any other node.

\[ \bigwedge\limits_{i,j,q,k} b_{i,j,q,k} \supset \neg \bigvee_{j \neq j^{\prime}, k^{\prime\prime}} ( a_{j^{\prime},k^{\prime\prime}} \land p_{k,k^{\prime\prime}}) \, \]

@@

D0: Only present edges can be dropped.
\[ \bigwedge\limits_{i,j,q} d_{i,j,q} \supset e_{i,j,q}  \]

@@

\[ \bigwedge\limits_{i,j}  [\bigvee_{q} (e_{i,j,q} \land  \neg d_{i,j,q}) \lor  (\bigvee_{i' != i}  r^{\prime}(i',j) \land  \bigvee_{q} (e_{i,i',q} \land \neg d_{i,i',q}) ] \supset r^{\prime}(i,j)  \]

>>>>

\begin{df}
  A VTS $G$ is tuple $(\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$, where
  \begin{itemize}
  \item $\nodes$ is a set of nodes representing compartments in the VTS,
  \item $\mols$ is the set of molecules flowing in the network, 
  \item $\nlabel : \nodes \maps \powerset{\mols}$ defines the molecules present in the nodes  
  \item $ \nodes \times (\powerset{\mols}-\emptyset) \times \nodes \subseteq \edges$ is the set of edges
    with molecule sets as labels
  \item $\pairs : \mols \times \mols$ is pairing relating
  \item $\nodef : \mols \maps \powerset{\mols} \maps \booleans$ is activity maps for nodes
  \item $\edgef : \mols \maps \powerset{\mols} \maps \booleans $ is activity maps for edges
  \end{itemize}
\end{df}
$\pairs$ defines which molecules can fuse with which molecules.
%
Let $\pairs(M')$ denote $\{m|(m,m') \in P \text{ and } m' \in M'\}$.
%
$\nodef$ and $\edgef$ are used to define activity of molecules on
nodes and edges.
%
A molecule $k$ is {\em active} at node $n$ if $k \in \nlabel(n)$ and
$\nodef(k,\nlabel(n))$ is true.
%
A molecule $k$ is {\em active} at edges $(n,M',n')$ if $k \in M'$ and
$\edgef(k,M')$ is true.
%
\ashu{define missing conds V1-V7}

@@

\item Boolean variable $n_{i,m}$ indicates if $m \in \nlabel(i)$
\item Boolean variable $e_{i,j,q}$ indicates if $q$th edge exists between $i$ and $j$.
\item Boolean variable $e_{i,j,q,m}$ indicates if $q$th edge between $i$ and $j$ contains molecule $m$.

\item Boolean variable $p_{k,k'}$ indicates if $(m,m') \in \pairs$
\item uninterpreted Boolean functions $\nodef_m : \booleans^\mu \maps \booleans$
encoding $\nodef(m)$ map
\item uninterpreted Boolean functions $\edgef_m : \booleans^\mu \maps \booleans$
encoding $\edgef(m)$ map
\end{enumerate}

We also have auxiliary variables that will help us encode the properties 
\begin{enumerate}
\item $a_{i,m}$ indicates that molecule $m$ is active at node $i$, i.e., $\nodef(m,L(i))$
  holds
\item $b_{i,j,q,m}$ indicates that molecule $m$ is active at $q$th edge $(i,M',j)$ between $i$ and $j$, i.e., $\edgef(m,M')$ holds
\item $r_{i,j,m,z}$ indicates that there is a path from $i$ to $j$ of length $z$
\item $d_{i,j,q}$ indicates $q$th edge between $i$ and $j$ is dropped(used in encoding
  $k$-connected property)
\end{enumerate}

% Every variable is a Boolean bit, for example, $n_{0,1}$ represents node 0
% with type 1 molecule present. Assignment to the Boolean variable
% n(0,1) to True will mean that at node 0 type 1 molecule is present.

% Basic notation for the formula we are trying to generate.
% \begin{enumerate}

% \item edge = e(i,j,q)      
%       we will assign e(i,j,q) for a qth edge between node i and j.

% \item dump = d(i,j,q)
%       to represent drop of edge(i,j,q) from the graph we will use d(i,j,q).

% \item presence node = n(i,k) 
% 	  kth molecule is present on the node i.

% \item active\_node = a(i,k) 
%       kth molecule is active on the node i.

% \item presence\_edge = e(i,j,q,k)
%       There is an qth edge between node i and j with kth molecule present.

% \item active\_edge = b(i,j,q,k)
%       There is an qth edge between node i and j with kth molecule active.

% \item reachability = r(i,j,k,z)
%       Node i and j are reachable with kth molecule present in z steps.

% \item pairing matrix = p(k,k')
%       Molecule k and k' are pairing molecules in the matrix.

% \item Type of the Boolean function = sorts (takes k-1 arguments Bool and return Bool) 
% \item Boolean function on nodes Function (fn\_{m},*sorts)

% \item Boolean function on edge Function(fe\_{m},*sorts)

% \end{enumerate}

We will describe the Boolean formulas that encode VTSs in several categories.
%
In the end, we will present in a table the formulas needed for the
model variants.
%
To avoid cumbersome notation, we will not explicitly write the ranges of the indexing
in the formulas.
%
$i$ and $j$ will be ranged over nodes.
%
$m$ will be ranged over molecules.
%
$q$ will be ranged over edges between two nodes.
%

% \subsubsection{Constraints on the structure of the edges, nodes and }
\subsubsection{Constraints on presence, activity of the molecule, and pairing matrix}
%
We need the following constraints~\eqref{eq:f0} and~\eqref{eq:c3}
to encode the basic structure of VTSs.
%
For an edge to exist it should have one molecule present. 
%
\begin{align}
  \bigwedge\limits_{i,j,q} (\bigvee_m e_{i,j,q,m} )\limplies e_{i,j,q}\tag{V1}\label{eq:f0}
\end{align}

If a molecule is active on an edge, it should be present on the edge.
%
\begin{align}
  \bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies e_{i,j,q,m}\tag{V2}\label{eq:f1}
\end{align}

A molecule should be present to be active on a node.  
\begin{align}
  \bigwedge\limits_{i,m} a_{i,m} \limplies n_{i,m}
  \tag{V3}\label{eq:c4}
\end{align}

The edge labels are subset of the node label of source and target compartments.

\begin{align}
  \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} \limplies n_{i,m}\tag{V4}\label{eq:c0}\\
  \bigwedge\limits_{i,j,q,m}  e_{i,j,q,m} \limplies n_{j,m}\tag{V5}\label{eq:c1}
\end{align}

%C1 =  e_ijk -> aik and e_ijk -> ajk

Self edges are not allowed.

\begin{align}
   \bigwedge\limits_{i,q} \neg e_{i,i,q}\tag{V6}\label{eq:c2}
\end{align}

Condition on $p_{kk'}$. \ashu{why???} Diagonal blocks should be all 0's.

\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V7}\label{eq:c3}
\end{align}

\subsubsection{Well-fused constraints}
Boolean constraint~\eqref{eq:fuse1} encodes that each edge must fuse with
destination node.
%
Boolean constraint~\eqref{eq:fuse2} encodes the each should not be able to
potentially fuse with any other node.

% Fusion rules consist of two different mechanisms.
% \begin{enumerate}
% \item  A \textbf{pairing mechanism} which determines compatible Q-R pairs on vesicles and compartments that can cause fusion.
% \item \textbf{regulatory mechanisms} on the edges and nodes (possibly distinct) which regulates molecules activity based on the presence/absence of other molecules on the corresponding node or edge.
% \end{enumerate}
% Boolean constraint~\eqref{eq:fuse1} and~\eqref{eq:fuse2} are used to encode well-fused
% property.

% F4: For an edge to be valid, at least one SNARE pair on the vesicle and target compartment must be active and have a non-zero entry in the pairing matrix.  
\begin{align}
  \bigwedge\limits_{i,j,q} e_{i,j,q} \limplies \bigvee_{m,m^{\prime}} (b_{i,j,q,m} \land a_{j,m^{\prime}} \land p_{m,m^{\prime}})
  \tag{V8}\label{eq:fuse1}  \\
\bigwedge\limits_{i,j,q,m} b_{i,j,q,m} \limplies \neg \bigvee_{j \neq j^{\prime}, m^{\prime\prime}} ( a_{j^{\prime},m^{\prime\prime}} \land p_{m,m^{\prime\prime}})
  \tag{V9}\label{eq:fuse2}  
\end{align}

% F5: To ensure that fusion respects the graph structure by the edge under consideration, it should not be possible to fuse with any other node.

\subsubsection{Regulation on nodes and edges}
%
Regulation of the molecules is based on the kind of the model under
consideration.
%
For example, model 1 (-V 1) has two different arbitrary
Boolean functions dictating the activity of the molecules on edge and
nodes.
%
For variation 3 (-V 3) activity on edge is driven by pairing
inhibition.

@@

A\_nn: No regulation on nodes i.e all the present molecules on the nodes are active.
\[
\bigwedge\limits_{i,m} n_{i,m} = a_{i,m} \,
\]

A\_nb: Boolean regulation on nodes. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,k} n_{i,m} \limplies a_{i,m} =   f_m (n_{i,1},\dots,n_{i,\mu}) \, \]

A\_en: No regulation on edges i.e all the present molecules on the edges are active.
\[ \bigwedge\limits_{i,j,q,m} e_{i,j,q,m} = b_{i,j,q,m} \, \]

A\_eb: Boolean regulation on edges. Activity of a molecule k on the node is defined as a Boolean function of presence of other molecule present on that node.
\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \limplies b_{i,j,q,k} == f_e[k](\bigvee_{k' != k} e_{i,j,q,k'}) \, \]

A\_ep: Pairing inhibition. The activity of the molecules is defined by inhibition by other molecules based on the pairing matrix. 
\[ \bigwedge\limits_{i,j,q, k}  (e_{i,j,q,k} \land [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \limplies e_{i,j,q,k'})] ) \limplies \neg b_{i,j,q,k}) \, \]

\[ \bigwedge\limits_{i,j, q, k}  (e_{i,j,q,k} \land \neg [ ( \bigvee_{k'} p_{k,k'} \land \bigwedge_{k' != k} p_{k,k'} \limplies e_{i,j,q,k'})] ) \limplies b_{i,j,q,k}) \, \]

% - F0: b(i,j,k) $\limplies$ e(i,j,k)

@@

% I have implemented: Changed the implementation in code too.
%\[ \bigwedge\limits_{i,j,q,k} e_{i,j,q,k} \limplies (r_{j,i,k,0}  \lor r_{j,i,k,1} \lor ... \lor r_{j,i,k,p}) \, \]

% Update: Better 
\[ \bigwedge\limits_{i,j,k} (\bigvee_{q} e_{i,j,q,k}) \limplies (r_{j,i,k,0}  \lor r_{j,i,k,1} \lor ... \lor r_{j,i,k,p}) \, \]

@@

\[ \bigwedge\limits_{i,j,k,p} r_{i,j,k,p} \limplies (\bigvee_{q} \, e_{i,j,q,k} \lor \bigvee_{i\neq i^{\prime}} ( \, \bigvee_{q} e_{i,i^{\prime},q,k}) \land r_{i^{\prime},j,k,p-1} ) \, \]

@@

D0: Only present edges can be dropped.
\[ \bigwedge\limits_{i,j,q} d_{i,j,q} \limplies e_{i,j,q}  \]

@@

\[ \bigwedge\limits_{i,j}  [\bigvee_{q} (e_{i,j,q} \land  \neg d_{i,j,q}) \lor  (\bigvee_{i' != i}  r^{\prime}(i',j) \land  \bigvee_{q} (e_{i,i',q} \land \neg d_{i,i',q}) ] \limplies r^{\prime}(i,j)  \]
